---
title: "Practical 2 Answers"
subtitle: "Working with Data: Data Sources and Data Manipulation"
author: "Your Name Here"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    link-external-newwindow: true
    link-external-icon: true
editor: source
---

<!--
Before you start:

- Make sure this file lives in the root of your project repository, alongside
  a `data/` folder containing:
    - reg-nurses-dataworld-1998-2020.csv
    - nurse-licenses-ncsbn-2026-01-05.csv
  (see the practical instructions for the download links).
- See the README.md in this repository for package installation instructions
  and Git/GitHub tips.
- Render this document with the "Render" button in RStudio, or by running
  quarto::quarto_render("practical2-answers.qmd") in the R console.
-->

## Getting started

```{r}
#| label: setup
#| message: false
#| warning: false

# Load the packages you need for this practical.
# See the README for installation instructions if any of these are missing.
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)
```

## US State Nurse Data

### Question 1

Download the `.csv` file and use `readr::read_csv()` to read the dataset into R.

```{r}
#| label: read-data

# nurses <- read_csv("data/reg-nurses-dataworld-1998-2020.csv")
```

### Question 2

Check the format of the data. Is it in a tidy format? Are there cleaning/tidying manipulations to do?

```{r}
#| label: inspect-data

```




## Data Manipulation and Visualisation

### Question 3

Summarise the data available in the dataset. What years/states/variables are present?

```{r}
#| label: summarise-data

```




### Question 4

Using `dplyr::filter()` and `ggplot()`, recreate the plot from the 'Getting Started' section of the practical brief.

```{r}
#| label: recreate-plot

```

### Question 5

Create a new plot to visualise some other information (of your choice) from the dataset.

```{r}
#| label: own-plot

```




## Adding additional data

### Question 6

Read in the additional dataset from the National Council of State Boards of Nursing.

```{r}
#| label: read-additional-data

# nurse_licenses <- read_csv("data/nurse-licenses-ncsbn-2026-01-05.csv")
```

### Question 7

Summarise the data available in this new dataset. What years/states/variables are present?

```{r}
#| label: summarise-additional-data

```




### Question 8

What data is overlapping between this new dataset and the original one? Think about the temporal overlap and which types of nurses are present in each dataset.




### Question 9

If it is appropriate, create a combined dataset using both sources of information. If not appropriate, explain why.

```{r}
#| label: combine-data

```




### Question 10

Create a visualisation (of your choice) from the new or combined dataset.

```{r}
#| label: plot-new-or-combined

```

## Merging additional data

### Question 11

Filter the `reg-nurses-dataworld-1998-2020` data to only include 2017.

```{r}
#| label: filter-2017

```

### Question 12

Load the `us_rent_income` data located in the `tidyr` package.

```{r}
#| label: load-rent-income

# data(us_rent_income, package = "tidyr")
```

### Question 13

How would it be possible to merge the two datasets?




### Question 14

Merge the two datasets using `dplyr::full_join()`.

```{r}
#| label: merge-rent-nurses

```

### Question 15

Visualise the relationship between rent and nurse wages in 2017 using the combined data.

```{r}
#| label: plot-rent-vs-wages

```
