---
title: "Practical 2 Answers"
subtitle: "Working with Data: Data Sources and Data Manipulation"
author: "Tia Light"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    link-external-newwindow: true
    link-external-icon: true
editor: source
---

<!--
Before you start:

- Make sure this file lives in the root of your project repository, alongside
  a `data/` folder containing:
    - reg-nurses-dataworld-1998-2020.csv
    - nurse-licenses-ncsbn-2026-01-05.csv
  (see the practical instructions for the download links).
- See the README.md in this repository for package installation instructions
  and Git/GitHub tips.
- Render this document with the "Render" button in RStudio, or by running
  quarto::quarto_render("practical2-answers.qmd") in the R console.
-->

## Getting started

```{r}
#| label: setup
#| message: false
#| warning: false

# Load the packages you need for this practical.
# See the README for installation instructions if any of these are missing.
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)
```

## US State Nurse Data

### Question 1

Download the `.csv` file and use `readr::read_csv()` to read the dataset into R.

```{r}
#| label: read-data

nurses <- read_csv("data/reg-nurses-dataworld-1998-2020.csv")
```

### Question 2

Check the format of the data. Is it in a tidy format? Are there cleaning/tidying manipulations to do?

```{r}
#| label: inspect-data

glimpse(nurses)

names(nurses)

# How much of each column is missing, and does the missingness cluster by
# year or by state?
nurses |>
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  glimpse()

nurses |> filter(is.na(`Location Quotient`)) |> count(Year)

nurses |> filter(is.na(`Total Employed RN`)) |> select(State, Year)


nurses <- nurses |>
  rename(
    total_employed        = `Total Employed RN`,
    employed_se_pct       = `Employed Standard Error (%)`,
    hourly_wage_avg       = `Hourly Wage Avg`,
    hourly_wage_median    = `Hourly Wage Median`,
    annual_salary_avg     = `Annual Salary Avg`,
    annual_salary_median  = `Annual Salary Median`,
    wage_se_pct           = `Wage/Salary standard error (%)`,
    hourly_p10            = `Hourly 10th Percentile`,
    hourly_p25            = `Hourly 25th Percentile`,
    hourly_p75            = `Hourly 75th Percentile`,
    hourly_p90            = `Hourly 90th Percentile`,
    annual_p10            = `Annual 10th Percentile`,
    annual_p25            = `Annual 25th Percentile`,
    annual_p75            = `Annual 75th Percentile`,
    annual_p90            = `Annual 90th Percentile`,
    location_quotient     = `Location Quotient`,
    national_total_employed          = `Total Employed (National)_Aggregate`,
    national_healthcare_employed     = `Total Employed (Healthcare, National)_Aggregate`,
    state_healthcare_employed        = `Total Employed (Healthcare, State)_Aggregate`,
    state_total_employed              = `Yearly Total Employed (State)_Aggregate`
  )

nurses

```




## Data Manipulation and Visualisation

### Question 3

Summarise the data available in the dataset. What years/states/variables are present?

```{r}
#| label: summarise-data

# column names
names(nurses)

# years of data
nurses |> summarise(min_year = min(Year),
                    max_year = max(Year),
                    n_years = n_distinct(Year),
                    n_states = n_distinct(State))

# check for each state (does each state have the same years?)
nurses |> 
  group_by(State) |>
  summarise(min_year = min(Year),
            max_year = max(Year),
            n_years = n_distinct(Year),
            n_rows = n())

# state names
state_names <- nurses |> pull(State) |> unique() |> sort()
state_names

# is Year x State a unique identified for the rows of the data?
nurses |> summarise(n_distinct(Year,State)) |> pull() == 54*23

```




### Question 4

Using `dplyr::filter()` and `ggplot()`, recreate the plot from the 'Getting Started' section of the practical brief.

```{r}
#| label: recreate-plot

 nurses |> filter(State %in% c('California', 'North Carolina', 'New York', 'Alabama')) |>
  ggplot() +
  geom_line(aes(y = hourly_wage_median, x = Year, colour = State)) +
  paletteer::scale_colour_paletteer_d("colorBlindness::paletteMartin") +
  scale_y_continuous("Hourly Wage Median") +
  theme_bw()

```

### Question 5

Create a new plot to visualise some other information (of your choice) from the dataset.

```{r}
#| label: own-plot
# How does pay vary across states in the most recent year available?
nurses |>
  filter(Year == max(Year), !is.na(annual_salary_median)) |>
  mutate(State = reorder(State, annual_salary_median)) |>
  ggplot(aes(x = annual_salary_median, y = State)) +
  geom_point() +
  scale_x_continuous("Median annual RN salary (US$)", labels = scales::label_dollar()) +
  ylab(NULL) +
  ggtitle("Median RN salary by state, 2020") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 6))
```




## Adding additional data

### Question 6

Read in the additional dataset from the National Council of State Boards of Nursing.

```{r}
#| label: read-additional-data

nurse_licenses <- read_csv("data/nurse-licenses-ncsbn-2026-01-05.csv")
```

### Question 7

Summarise the data available in this new dataset. What years/states/variables are present?

```{r}
#| label: summarise-additional-data

nrow(nurse_licenses) # not 54, why?

nurse_licenses$Jurisdiction

nurse_licenses |> filter(grepl("-", Jurisdiction))

nurse_licenses_clean <- nurse_licenses |>
  # split Jurisdiction column on "-" character
  separate_wider_delim(Jurisdiction, delim = "-", names = c("State","Nurse"), too_few = "align_start") |>
  # replace NAs with "All"
  mutate(Nurse = ifelse(is.na(Nurse), "All", Nurse)) |> 
  # change state name to lower case
  mutate(state_lwc = stringr::str_to_lower(State)) |>
  # remove State column
  select(-State)

# sum up states that have "split" records of nurses i.e. RN + PN + VN
nurse_licenses_all <- nurse_licenses_clean |> 
  group_by(state_lwc) |>
  summarise(licenses_2026 = sum(Number))

nurse_licenses_all
```




### Question 8

What data is overlapping between this new dataset and the original one? Think about the temporal overlap and which types of nurses are present in each dataset.

Temporal overlap: none. nurses runs 1998–2020. The nurse_licenses data is a snapshot from January 2026 (that Josh downloaded). This is six years after the other dataset ends, and a single moment in time rather than a year. There is no year we can match exactly between the two.

Conceptual overlap: partial, and not a clean match. The two datasets measure different things:

nurses (Total Employed RN) is BLS OES employment data: an estimate of how many RNs were actively working in a state that year, from surveying employers.
nurse_licenses (Number) is NCSBN licensure data: a count of active licenses held, from state licensing boards. A license count will generally be higher than employment (retired-but-licensed, between jobs, working out of state on a compact license, etc.). We also summed the types of nurse registration PN + VN + RN.
Geographically the overlap is better: both datasets use full state/ territory names (nurses in Title Case, nurse_licenses in ALL CAPS, so case-normalisation is needed), and both include some of the same territories, though nurse_licenses additionally includes American Samoa and the Northern Mariana Islands, which nurses doesn’t have at all.


### Question 9

If it is appropriate, create a combined dataset using both sources of information. If not appropriate, explain why.

```{r}
#| label: combine-data

## Given the mismatches above, a direct, literal merge would be misleading: there’s no shared year to join on, and naively joining on state name alone would implicitly compare 2020 employment against a 2026 license count as if they were the same point in time and the same population of nurses. We don’t think that’s appropriate to present as a genuine “combined dataset” without heavy caveats.

## What is reasonable is a clearly-labelled data comparison — e.g. “how does the most recent employment figure we have compare to the most recent license count we have” — as long as we’re explicit that the two numbers are years apart and measure different underlying populations. We do this below.

employed_nurses_2020 <- nurses |>
  filter(Year == 2020) |>
  select(State, total_employed) |>
  rename(employed_2020 = total_employed) |> 
  mutate(state_lwc = stringr::str_to_lower(State))

employed_vs_licenses <- employed_nurses_2020 |>
  inner_join(nurse_licenses_all, by = "state_lwc") |>
  select(-state_lwc) 

```




### Question 10

Create a visualisation (of your choice) from the new or combined dataset.

```{r}
#| label: plot-new-or-combined

ggplot(employed_vs_licenses, aes(x = employed_2020, y = licenses_2026)) + 
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_bw()
```

## Merging additional data

### Question 11

Filter the `reg-nurses-dataworld-1998-2020` data to only include 2017.

```{r}
#| label: filter-2017

nurses_2017 <- nurses |>
  filter(Year == 2017)

nurses_2017
```

### Question 12

Load the `us_rent_income` data located in the `tidyr` package.

```{r}
#| label: load-rent-income

data(us_rent_income, package = "tidyr")
us_rent_income

```

### Question 13

How would it be possible to merge the two datasets?

How would it be possible to merge the two datasets?

us_rent_income is in long format: for each state NAME (or GEOID) there are two rows, one with variable == "income" and one with variable == "rent", each with an estimate and a margin of error (moe) — 104 rows in total (52 states/territories x 2 variables). To join it to nurses_2017 (one row per state) by state, we either need to

pivot it wider so that income and rent each become their own column (one row per state) … otherwise a join would duplicate every nurses_2017 row once for income and once for rent.

Filter down to income or rent (assuming we are only interested in one).

We will use the first option.



### Question 14

Merge the two datasets using `dplyr::full_join()`.

```{r}
#| label: merge-rent-nurses

us_rent_income_wide <- us_rent_income |>
  select(-moe) |>
  pivot_wider(names_from = variable, values_from = estimate)

rent_nurses_2017 <- full_join(nurses_2017, us_rent_income_wide, by = c("State" = "NAME"))

rent_nurses_2017

```

### Question 15

Visualise the relationship between rent and nurse wages in 2017 using the combined data.

```{r}
#| label: plot-rent-vs-wages

rent_nurses_2017 |>
  filter(!is.na(rent), !is.na(annual_salary_median)) |>
  ggplot(aes(x = rent, y = annual_salary_median)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggrepel::geom_text_repel(aes(label = State), size = 2.5) +
  scale_x_continuous("Median monthly rent (US$)", labels = scales::label_dollar()) +
  scale_y_continuous("Median annual RN salary (US$)", labels = scales::label_dollar()) +
  ggtitle("RN salary vs. median rent by state, 2017") +
  theme_bw()


```
